---
title: "01 — Introduction to NIR Spectroscopy and the R Environment"
subtitle: "CAC 2026 · Open-source chemometrics for real-world NIR handheld spectroscopy"
author: "Leonardo Ramirez-Lopez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 4.5
)
```

## 1.1 What is NIR spectroscopy?

Near-infrared (NIR) spectroscopy measures the interaction of light with matter in
the wavelength range of approximately **700–2500 nm**.  When a sample is illuminated
by NIR radiation, the molecules absorb energy at frequencies that correspond to
overtones and combinations of fundamental molecular vibrations — mainly C–H, O–H,
and N–H bonds.

Because organic molecules absorb NIR radiation in characteristic (though overlapping)
patterns, NIR spectra carry rich information about chemical composition.  However,
extracting quantitative information requires **chemometric** (multivariate statistical)
methods — the focus of this course.

### Why handheld NIR?

Modern portable and handheld NIR instruments enable:

* **In-field** measurements — no sample preparation, no laboratory, no waiting.
* **High throughput** — hundreds of samples per day per operator.
* **Non-destructive** analysis — the sample is preserved after measurement.

The trade-off is that handheld instruments typically have lower signal-to-noise ratios
and are affected by ambient conditions (temperature, orientation, surface roughness).
Robust chemometric workflows are therefore essential.

---

## 1.2 The R environment for chemometrics

R provides a mature, open-source ecosystem for spectroscopic data analysis.  The
packages used in this course are summarised below.

```{r packages}
library(prospectr)  # spectral pre-processing and sample selection
library(resemble)   # memory-based learning
library(pls)        # PLS and PCR regression
library(ggplot2)    # graphics
library(dplyr)      # data manipulation
library(tidyr)      # data reshaping
```

### Key packages at a glance

| Package | Main functions |
|---------|---------------|
| `prospectr` | `standardNormalVariate()`, `msc()`, `savitzkyGolay()`, `kenStone()` |
| `resemble` | `orthoProjection()`, `mbl()` |
| `pls` | `plsr()`, `pcr()`, `crossval()` |
| `ggplot2` | `ggplot()`, `geom_line()`, `geom_point()` |

---

## 1.3 Checking your installation

Run the following to confirm that everything is set up correctly.

```{r check-install}
pkgs <- c("prospectr", "resemble", "pls", "ggplot2", "dplyr", "tidyr", "caret")
status <- sapply(pkgs, requireNamespace, quietly = TRUE)
data.frame(
  Package   = names(status),
  Installed = status
)
```

If any package shows `FALSE`, rerun `source("../install_packages.R")` from the
project root.

---

## 1.4 The analytical workflow

The diagram below outlines the steps covered in this course:

```
Raw NIR spectra
      │
      ▼
Pre-processing          (Notebook 03)
  • SNV / MSC
  • Smoothing & derivatives
      │
      ▼
Exploratory analysis    (Notebook 04)
  • PCA scores & loadings
  • Outlier detection
      │
      ▼
Calibration             (Notebook 05)
  • PLS regression
  • Number of components selection
      │
      ▼
Validation              (Notebook 06)
  • Cross-validation
  • External test set
  • Prediction uncertainty
```

Each step will be demonstrated with real NIR soil data from the `prospectr`
package's built-in `NIRsoil` dataset.

---

## 1.5 First look at the NIRsoil dataset

```{r load-nirsoil}
data(NIRsoil)

# NIRsoil is a list; the spectra are in $spc
dim(NIRsoil$spc)   # samples × wavelengths
head(NIRsoil$spc[, 1:5])

# Reference data
head(NIRsoil[c("Nt", "Ciso", "CEC", "moisture")])
```

```{r summary-reference}
summary(NIRsoil[c("Nt", "Ciso", "CEC", "moisture")])
```

The dataset contains **`r nrow(NIRsoil$spc)` samples** measured at
**`r ncol(NIRsoil$spc)` wavelengths** (1100–2498 nm, 2 nm steps).  Four
reference properties are available; we will focus on **total nitrogen (`Nt`)**
throughout the course.

---

## 1.6 Session information

```{r session-info}
sessionInfo()
```
