---
title: "02 — Data Loading and Spectral Visualisation"
subtitle: "CAC 2026 · Open-source chemometrics for real-world NIR handheld spectroscopy"
author: "Leonardo Ramirez-Lopez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 5
)
```

```{r packages}
library(prospectr)
library(ggplot2)
library(dplyr)
library(tidyr)
```

---

## 2.1 Loading spectral data

### From the built-in `NIRsoil` dataset

```{r load-data}
data(NIRsoil)

# Spectra matrix: rows = samples, columns = wavelengths
spectra  <- NIRsoil$spc
ref_data <- NIRsoil[c("Nt", "Ciso", "CEC", "moisture")]

wavelengths <- as.numeric(colnames(spectra))
cat("Wavelength range:", range(wavelengths), "nm\n")
cat("Number of samples:", nrow(spectra), "\n")
```

### Loading your own CSV data

If you have spectra exported from a handheld instrument as a CSV file, use:

```{r load-csv, eval=FALSE}
# Adjust the path and column names to match your file
raw <- read.csv("../data/my_spectra.csv", check.names = FALSE)

# Identify which columns are wavelengths (numeric column names)
wl_cols   <- grep("^[0-9]", names(raw), value = TRUE)
meta_cols <- setdiff(names(raw), wl_cols)

spectra     <- as.matrix(raw[, wl_cols])
wavelengths <- as.numeric(wl_cols)
ref_data    <- raw[, meta_cols, drop = FALSE]
```

---

## 2.2 Quick data inspection

```{r inspect}
# Check for missing values
cat("Missing values in spectra:", sum(is.na(spectra)), "\n")
cat("Missing values in Nt:     ", sum(is.na(ref_data$Nt)), "\n")

# Absorbance range
cat("Absorbance range:", round(range(spectra), 3), "\n")
```

---

## 2.3 Spectral visualisation

### Plot all spectra

```{r plot-all-spectra}
# Convert to long format for ggplot2
spectra_long <- spectra |>
  as.data.frame() |>
  tibble::rownames_to_column("sample") |>
  tidyr::pivot_longer(-sample, names_to = "wavelength", values_to = "absorbance") |>
  dplyr::mutate(wavelength = as.numeric(wavelength))

ggplot(spectra_long, aes(x = wavelength, y = absorbance, group = sample)) +
  geom_line(alpha = 0.15, colour = "#2166AC") +
  labs(
    title = "NIRsoil — raw absorbance spectra",
    x     = "Wavelength (nm)",
    y     = "Absorbance"
  ) +
  theme_bw()
```

### Colour spectra by reference value

Colouring by a reference property reveals whether spectral variation is related
to the analyte of interest.

```{r plot-coloured}
spectra_long_nt <- spectra |>
  as.data.frame() |>
  tibble::rownames_to_column("sample") |>
  dplyr::mutate(Nt = ref_data$Nt) |>
  tidyr::pivot_longer(-c(sample, Nt),
                      names_to  = "wavelength",
                      values_to = "absorbance") |>
  dplyr::mutate(wavelength = as.numeric(wavelength)) |>
  dplyr::filter(!is.na(Nt))

ggplot(spectra_long_nt, aes(x = wavelength, y = absorbance,
                             group = sample, colour = Nt)) +
  geom_line(alpha = 0.4) +
  scale_colour_viridis_c(option = "plasma", name = "Nt (%)") +
  labs(
    title = "NIRsoil spectra coloured by total nitrogen (Nt)",
    x     = "Wavelength (nm)",
    y     = "Absorbance"
  ) +
  theme_bw()
```

---

## 2.4 Distribution of the reference property

```{r hist-nt}
nt_df <- data.frame(Nt = ref_data$Nt) |> dplyr::filter(!is.na(Nt))

ggplot(nt_df, aes(x = Nt)) +
  geom_histogram(bins = 30, fill = "#2166AC", colour = "white") +
  geom_vline(xintercept = mean(nt_df$Nt), linetype = "dashed",
             colour = "tomato", linewidth = 0.8) +
  labs(
    title = "Distribution of total nitrogen (Nt)",
    x     = "Nt (%)",
    y     = "Count"
  ) +
  theme_bw()
```

```{r summary-nt}
summary(nt_df$Nt)
```

---

## 2.5 Single-sample spectrum

```{r single-spectrum}
sample_idx <- 1
single_df  <- data.frame(
  wavelength = wavelengths,
  absorbance = spectra[sample_idx, ]
)

ggplot(single_df, aes(x = wavelength, y = absorbance)) +
  geom_line(colour = "#1B7837", linewidth = 0.8) +
  annotate("rect", xmin = 1400, xmax = 1450,
           ymin = -Inf, ymax = Inf, alpha = 0.15, fill = "orange") +
  annotate("rect", xmin = 1900, xmax = 1960,
           ymin = -Inf, ymax = Inf, alpha = 0.15, fill = "orange") +
  annotate("text", x = 1425, y = max(single_df$absorbance) * 0.95,
           label = "O–H", colour = "darkorange", size = 3) +
  annotate("text", x = 1930, y = max(single_df$absorbance) * 0.95,
           label = "O–H", colour = "darkorange", size = 3) +
  labs(
    title   = paste("Single NIR spectrum — sample", sample_idx),
    subtitle = "Shaded regions: water absorption bands",
    x = "Wavelength (nm)",
    y = "Absorbance"
  ) +
  theme_bw()
```

---

## 2.6 Summary

In this notebook we have:

* Loaded the `NIRsoil` dataset from the `prospectr` package.
* Visualised the full spectral stack and related spectral features to the
  analyte of interest.
* Examined the distribution of the reference property.

In **Notebook 03** we will apply spectral pre-processing to reduce physical
variation unrelated to chemical composition.
