---
title: "03 — Spectral Pre-processing"
subtitle: "CAC 2026 · Open-source chemometrics for real-world NIR handheld spectroscopy"
author: "Leonardo Ramirez-Lopez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 5
)
```

```{r packages}
library(prospectr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
```

```{r load-data}
data(NIRsoil)
spectra     <- NIRsoil$spc
wavelengths <- as.numeric(colnames(spectra))
```

---

## 3.1 Why pre-process NIR spectra?

Raw NIR spectra collected in the field are affected by **physical** sources of
variation that are unrelated to the analyte of interest:

| Effect | Cause | Correction |
|--------|-------|------------|
| Additive baseline shifts | Variable path length, particle size | SNV, de-trending |
| Multiplicative scaling | Reflectance differences, instrument temperature | SNV, MSC |
| High-frequency noise | Electronic noise | Savitzky–Golay smoothing |
| Overlapping broad bands | Instrument optics, water vapour | Derivatives |

Pre-processing reduces these effects and often improves calibration model
performance.

---

## 3.2 Standard Normal Variate (SNV)

SNV centres each individual spectrum to zero mean and scales it to unit variance.
It corrects **both** additive and multiplicative scatter effects.

$$\text{SNV}(x_{i,j}) = \frac{x_{i,j} - \bar{x}_i}{s_i}$$

where $\bar{x}_i$ and $s_i$ are the mean and standard deviation of the $i$-th
spectrum.

```{r snv}
spectra_snv <- standardNormalVariate(spectra)
```

```{r plot-snv, echo=FALSE}
make_long <- function(mat, wl, label) {
  as.data.frame(mat) |>
    tibble::rownames_to_column("sample") |>
    tidyr::pivot_longer(-sample,
                        names_to  = "wavelength",
                        values_to = "value") |>
    dplyr::mutate(wavelength = as.numeric(wavelength),
                  transform  = label)
}

raw_long <- make_long(spectra,     wavelengths, "Raw")
snv_long <- make_long(spectra_snv, wavelengths, "SNV")

bind_rows(raw_long, snv_long) |>
  ggplot(aes(x = wavelength, y = value, group = sample)) +
  geom_line(alpha = 0.15, colour = "#2166AC") +
  facet_wrap(~transform, scales = "free_y") +
  labs(x = "Wavelength (nm)", y = "Value",
       title = "Raw vs SNV-corrected spectra") +
  theme_bw()
```

---

## 3.3 Multiplicative Scatter Correction (MSC)

MSC regresses each spectrum against a reference spectrum (usually the mean
spectrum) and uses the regression coefficients to remove additive and
multiplicative effects.

```{r msc}
spectra_msc <- msc(spectra)
```

```{r plot-msc, echo=FALSE}
msc_long <- make_long(spectra_msc, wavelengths, "MSC")

bind_rows(raw_long, msc_long) |>
  ggplot(aes(x = wavelength, y = value, group = sample)) +
  geom_line(alpha = 0.15, colour = "#1B7837") +
  facet_wrap(~transform, scales = "free_y") +
  labs(x = "Wavelength (nm)", y = "Value",
       title = "Raw vs MSC-corrected spectra") +
  theme_bw()
```

---

## 3.4 Savitzky–Golay smoothing and derivatives

The Savitzky–Golay (SG) filter fits a polynomial to a moving window of
consecutive spectral points.  Setting the derivative order > 0 yields the
first or second derivative of the spectrum, which:

* Removes additive baseline offsets (1st derivative).
* Removes both additive and multiplicative offsets (2nd derivative).
* Sharpens spectral features and highlights peak positions.

Key parameters:

| Parameter | Argument | Typical values |
|-----------|----------|----------------|
| Window width | `w` | 5–21 (odd number) |
| Polynomial order | `p` | 2–4 |
| Derivative order | `m` | 0 (smooth), 1, 2 |

```{r sg-1st}
# 1st derivative: window = 11, polynomial order = 2
spectra_sg1 <- savitzkyGolay(spectra, m = 1, p = 2, w = 11)
```

```{r sg-2nd}
# 2nd derivative: window = 11, polynomial order = 2
spectra_sg2 <- savitzkyGolay(spectra, m = 2, p = 2, w = 11)
```

```{r plot-sg, echo=FALSE}
wl_sg1 <- as.numeric(colnames(spectra_sg1))
wl_sg2 <- as.numeric(colnames(spectra_sg2))

sg1_long <- make_long(spectra_sg1, wl_sg1, "SG 1st deriv.")
sg2_long <- make_long(spectra_sg2, wl_sg2, "SG 2nd deriv.")

bind_rows(raw_long, sg1_long, sg2_long) |>
  ggplot(aes(x = wavelength, y = value, group = sample)) +
  geom_line(alpha = 0.15, colour = "#762A83") +
  facet_wrap(~transform, scales = "free_y", ncol = 1) +
  labs(x = "Wavelength (nm)", y = "Value",
       title = "Raw and Savitzky–Golay derivatives") +
  theme_bw(base_size = 11)
```

---

## 3.5 Comparing pre-processing methods

The table below summarises the properties corrected by each method:

| Method | Additive shift | Multiplicative | Noise | Derivative |
|--------|:--------------:|:--------------:|:-----:|:----------:|
| SNV | ✓ | ✓ | ✗ | ✗ |
| MSC | ✓ | ✓ | ✗ | ✗ |
| SG (m=0) | ✗ | ✗ | ✓ | ✗ |
| SG 1st deriv. | ✓ | ✗ | partial | ✓ |
| SG 2nd deriv. | ✓ | ✓ | partial | ✓ |

In practice, combinations are common:

```{r combined}
# SNV followed by 1st SG derivative (a very common pipeline)
spectra_snv_sg1 <- savitzkyGolay(standardNormalVariate(spectra),
                                  m = 1, p = 2, w = 11)
dim(spectra_snv_sg1)
```

---

## 3.6 Choosing a pre-processing method

There is no universally best method — the choice depends on:

1. **The instrument**: diffuse reflectance vs. transmittance; single-beam vs.
   dual-beam.
2. **The sample matrix**: powders, liquids, intact fruits, soils, etc.
3. **The analyte**: strong absorbers may not need derivatives; weak absorbers
   often benefit.
4. **Model performance**: compare RMSECV on a representative calibration set
   using several pre-processing options.

We will select a pre-processing method in Notebook 05 based on cross-validation
error.

---

## 3.7 Summary

We applied three standard pre-processing methods using `prospectr`:

* `standardNormalVariate()` — SNV
* `msc()` — MSC
* `savitzkyGolay()` — smoothing and derivatives

In **Notebook 04** we use the pre-processed spectra for exploratory analysis
with PCA.
