---
title: "04 — Exploratory Data Analysis with PCA"
subtitle: "CAC 2026 · Open-source chemometrics for real-world NIR handheld spectroscopy"
author: "Leonardo Ramirez-Lopez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 5.5
)
```

```{r packages}
library(prospectr)
library(resemble)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
```

```{r load-data}
data(NIRsoil)
spectra     <- NIRsoil$spc
ref_data    <- NIRsoil[c("Nt", "Ciso", "CEC", "moisture")]
wavelengths <- as.numeric(colnames(spectra))

# Apply SNV pre-processing
spectra_snv <- standardNormalVariate(spectra)
```

---

## 4.1 Why PCA?

Principal component analysis (PCA) projects the high-dimensional spectral matrix
(**n** samples × **p** wavelengths) into a low-dimensional space of **principal
components** (PCs) that capture maximum variance.

Benefits for spectroscopy:

* **Visualise** the overall spread and structure of the dataset.
* **Detect outliers** — unusual samples that may be mislabelled or contaminated.
* **Understand** which wavelength regions drive the main sources of variation.
* **Compress** spectra into a compact set of scores for subsequent modelling.

---

## 4.2 Computing PCA with `resemble`

The `resemble` package provides the `orthoProjection()` function for efficient
PCA and PLS projection of spectral data.

```{r pca-resemble}
pca_result <- orthoProjection(
  Xr = spectra_snv,    # input matrix (pre-processed spectra)
  pc = 20,             # maximum number of components to extract
  method = "pca",      # principal component analysis
  center = TRUE,
  scale  = FALSE
)

# Variance explained
var_exp <- pca_result$variance$x_var
cumvar  <- cumsum(var_exp)

var_df <- data.frame(
  PC      = seq_along(var_exp),
  VarExp  = var_exp * 100,
  CumVar  = cumvar * 100
)

head(var_df, 10)
```

```{r scree-plot}
ggplot(var_df[1:15, ], aes(x = PC)) +
  geom_col(aes(y = VarExp), fill = "#2166AC", alpha = 0.7) +
  geom_line(aes(y = CumVar), colour = "tomato", linewidth = 0.8) +
  geom_point(aes(y = CumVar), colour = "tomato", size = 2) +
  scale_y_continuous(
    name     = "Variance explained (%)",
    sec.axis = sec_axis(~ ., name = "Cumulative variance (%)")
  ) +
  scale_x_continuous(breaks = 1:15) +
  labs(title = "PCA scree plot — NIRsoil (SNV)", x = "Principal component") +
  theme_bw()
```

> **Rule of thumb**: retain enough PCs to explain ≥ 99% of spectral variance for
> exploratory purposes.  For calibration, the optimal number of PLS components
> is selected by cross-validation (Notebook 05).

---

## 4.3 Score plots

### PC1 vs PC2

```{r scores-pc12}
scores <- as.data.frame(pca_result$scores)
scores$Nt <- ref_data$Nt

ggplot(scores, aes(x = PC1, y = PC2, colour = Nt)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_colour_viridis_c(option = "plasma", name = "Nt (%)") +
  labs(
    title = "PCA score plot — PC1 vs PC2",
    subtitle = paste0(
      "PC1: ", round(var_df$VarExp[1], 1), "% | ",
      "PC2: ", round(var_df$VarExp[2], 1), "% variance explained"
    )
  ) +
  theme_bw()
```

Samples are coloured by total nitrogen.  A gradient along the principal
components suggests that spectral variation is correlated with Nt.

### PC1 vs PC3

```{r scores-pc13}
ggplot(scores, aes(x = PC1, y = PC3, colour = Nt)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_colour_viridis_c(option = "plasma", name = "Nt (%)") +
  labs(
    title = "PCA score plot — PC1 vs PC3",
    subtitle = paste0(
      "PC1: ", round(var_df$VarExp[1], 1), "% | ",
      "PC3: ", round(var_df$VarExp[3], 1), "% variance explained"
    )
  ) +
  theme_bw()
```

---

## 4.4 Loadings plots

PCA loadings reveal which wavelengths contribute most to each component.

```{r loadings}
loadings <- as.data.frame(pca_result$loadings)
loadings$wavelength <- wavelengths

p1 <- ggplot(loadings, aes(x = wavelength, y = PC1)) +
  geom_line(colour = "#1B7837") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Wavelength (nm)", y = "Loading", title = "PC1 loadings") +
  theme_bw()

p2 <- ggplot(loadings, aes(x = wavelength, y = PC2)) +
  geom_line(colour = "#762A83") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Wavelength (nm)", y = "Loading", title = "PC2 loadings") +
  theme_bw()

p1 / p2
```

---

## 4.5 Outlier detection

### Hotelling's T² and Q residuals

Two statistics are commonly used to identify spectral outliers:

* **Hotelling's T²** measures how far a sample is from the centre of the
  multivariate space spanned by the retained PCs.
* **Q residual** (also called SPE, Squared Prediction Error) measures how well
  the retained PCs reconstruct the original spectrum — samples with high Q are
  not well described by the model.

```{r outliers}
n_pc <- 5  # number of PCs to use for outlier detection

# T² statistic
scores_5pc <- as.matrix(scores[, paste0("PC", 1:n_pc)])
T2 <- apply(scores_5pc, 1, function(x) {
  lambda <- var_exp[1:n_pc] * (nrow(spectra_snv) - 1)
  sum(x^2 / lambda)
})

# Q residuals (reconstruction error)
recon   <- scores_5pc %*% t(pca_result$loadings[, 1:n_pc])
Q_resid <- rowSums((spectra_snv - recon)^2)

# Thresholds (95th percentile)
T2_thresh <- quantile(T2,     0.95)
Q_thresh  <- quantile(Q_resid, 0.95)

outlier_df <- data.frame(
  sample    = seq_len(nrow(spectra_snv)),
  T2        = T2,
  Q         = Q_resid,
  outlier_T2 = T2 > T2_thresh,
  outlier_Q  = Q_resid > Q_thresh
)

ggplot(outlier_df, aes(x = T2, y = Q,
                        colour = outlier_T2 | outlier_Q)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_vline(xintercept = T2_thresh, linetype = "dashed", colour = "tomato") +
  geom_hline(yintercept = Q_thresh,  linetype = "dashed", colour = "tomato") +
  scale_colour_manual(values = c("FALSE" = "#2166AC", "TRUE" = "tomato"),
                      name = "Outlier") +
  labs(
    title = "Outlier detection — Hotelling T² vs Q residuals",
    x = "Hotelling T²", y = "Q residual (SPE)"
  ) +
  theme_bw()
```

```{r outlier-count}
cat("Samples flagged by T² (> 95th percentile):", sum(outlier_df$outlier_T2), "\n")
cat("Samples flagged by Q  (> 95th percentile):", sum(outlier_df$outlier_Q),  "\n")
```

> ⚠️ Outlier detection is exploratory.  A flagged sample should be investigated
> (was it measured incorrectly?  is it a genuinely unusual sample that expands
> the calibration range?) — not automatically removed.

---

## 4.6 Summary

We used `resemble::orthoProjection()` for PCA and explored:

* Variance explained (scree plot).
* Score plots coloured by the reference property.
* Loadings to interpret which wavelengths drive each component.
* Simple outlier detection via Hotelling's T² and Q residuals.

In **Notebook 05** we build a PLS regression calibration model.
