---
title: "05 — PLS Calibration"
subtitle: "CAC 2026 · Open-source chemometrics for real-world NIR handheld spectroscopy"
author: "Leonardo Ramirez-Lopez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 5
)
```

```{r packages}
library(prospectr)
library(pls)
library(ggplot2)
library(dplyr)
```

---

## 5.1 Partial Least Squares (PLS) regression

PLS regression (also called PLSR or PLS1 for a single response) decomposes
both the spectral matrix **X** and the response vector **y** into latent
variables (LVs) that maximise the covariance between **X** and **y**.

Compared with PCA + MLR, PLS has key advantages for NIR calibration:

* Latent variables are supervised — they are oriented toward predicting **y**.
* It is robust when $p \gg n$ (more wavelengths than samples).
* It naturally handles collinear predictors.

---

## 5.2 Data preparation

```{r data-prep}
data(NIRsoil)

spectra     <- NIRsoil$spc
wavelengths <- as.numeric(colnames(spectra))

# Pre-processing: SNV
spectra_snv <- standardNormalVariate(spectra)

# Remove samples with missing Nt
keep <- !is.na(NIRsoil$Nt)
X    <- spectra_snv[keep, ]
y    <- NIRsoil$Nt[keep]

cat("Samples for modelling:", nrow(X), "\n")
cat("Nt range:", round(range(y), 3), "%\n")
```

### Train / test split using Kennard–Stone

We use the `kenStone()` function from `prospectr` to select a representative
calibration set via the Kennard–Stone algorithm, ensuring the calibration
samples span the spectral space.

```{r train-test-split}
set.seed(42)
ks <- kenStone(X, k = round(nrow(X) * 0.7), metric = "mahal", pc = 10)

X_cal  <- X[ks$model, ]
y_cal  <- y[ks$model]
X_test <- X[ks$test,  ]
y_test <- y[ks$test]

cat("Calibration samples:", nrow(X_cal), "\n")
cat("Test samples:       ", nrow(X_test), "\n")
```

---

## 5.3 Fitting a PLS model

```{r pls-fit}
# pls::plsr expects the response as a matrix
pls_model <- plsr(
  y_cal ~ X_cal,
  ncomp      = 20,
  validation = "CV",       # leave-one-out cross-validation
  segments   = 10,         # 10-fold CV (faster than LOO for large datasets)
  method     = "oscorespls"
)
```

---

## 5.4 Selecting the optimal number of components

```{r rmsecv-plot}
rmsecv <- RMSEP(pls_model, estimate = "CV")$val[1, 1, ]
ncomp_seq <- 0:(length(rmsecv) - 1)

rmsecv_df <- data.frame(ncomp = ncomp_seq, RMSECV = rmsecv)

opt_ncomp <- ncomp_seq[which.min(rmsecv)]
cat("Optimal number of components (minimum RMSECV):", opt_ncomp, "\n")

ggplot(rmsecv_df, aes(x = ncomp, y = RMSECV)) +
  geom_line(colour = "#2166AC") +
  geom_point(colour = "#2166AC", size = 2) +
  geom_vline(xintercept = opt_ncomp, linetype = "dashed", colour = "tomato") +
  annotate("text", x = opt_ncomp + 0.3,
           y    = max(rmsecv_df$RMSECV) * 0.97,
           label = paste("Optimal:", opt_ncomp, "LVs"),
           colour = "tomato", hjust = 0, size = 3.5) +
  labs(
    title = "PLS cross-validation — RMSECV vs number of components",
    x     = "Number of latent variables",
    y     = "RMSECV (%)"
  ) +
  theme_bw()
```

> **Parsimony principle**: prefer the simplest model (fewest LVs) whose RMSECV
> is within one standard error of the minimum.  This reduces the risk of
> over-fitting.

---

## 5.5 Calibration performance

```{r cal-performance}
# Calibration predictions
y_hat_cal <- predict(pls_model, ncomp = opt_ncomp)[, , 1]

cal_df <- data.frame(
  measured  = y_cal,
  predicted = y_hat_cal
)

# Calibration statistics
rmse_cal <- sqrt(mean((cal_df$measured - cal_df$predicted)^2))
r2_cal   <- cor(cal_df$measured, cal_df$predicted)^2

cat(sprintf("Calibration — RMSEC: %.4f | R²: %.4f\n", rmse_cal, r2_cal))
```

```{r plot-cal}
ggplot(cal_df, aes(x = measured, y = predicted)) +
  geom_point(alpha = 0.6, colour = "#2166AC") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title    = paste0("Calibration — ", opt_ncomp, " LVs"),
    subtitle = sprintf("RMSEC = %.4f | R² = %.4f", rmse_cal, r2_cal),
    x = "Measured Nt (%)",
    y = "Predicted Nt (%)"
  ) +
  coord_equal() +
  theme_bw()
```

---

## 5.6 PLS regression coefficients

PLS regression coefficients (in the original wavelength space) reveal which
spectral regions are important for predicting Nt.

```{r reg-coeff}
coef_pls <- as.numeric(coef(pls_model, ncomp = opt_ncomp, intercept = FALSE))

coef_df <- data.frame(
  wavelength = wavelengths,
  coefficient = coef_pls
)

ggplot(coef_df, aes(x = wavelength, y = coefficient)) +
  geom_line(colour = "#762A83") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = paste0("PLS regression coefficients — ", opt_ncomp, " LVs"),
    x     = "Wavelength (nm)",
    y     = "Regression coefficient"
  ) +
  theme_bw()
```

Large positive or negative coefficients indicate wavelengths most influential
for predicting total nitrogen.

---

## 5.7 Summary

We:

* Split the dataset into calibration and test sets using Kennard–Stone.
* Fitted a PLS regression model with cross-validation.
* Selected the optimal number of latent variables by minimising RMSECV.
* Visualised calibration fit and regression coefficients.

In **Notebook 06** we evaluate the model on the independent test set and
compute full validation statistics.
