---
title: "06 — Model Validation and Diagnostics"
subtitle: "CAC 2026 · Open-source chemometrics for real-world NIR handheld spectroscopy"
author: "Leonardo Ramirez-Lopez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 5
)
```

```{r packages}
library(prospectr)
library(pls)
library(ggplot2)
library(dplyr)
library(patchwork)
```

---

## 6.1 Why validate?

A good calibration **RMSEC** alone is insufficient evidence of a useful model.
A model can over-fit the calibration data, performing well on calibration samples
but poorly on new, unseen samples.

**Validation** on an independent test set provides an unbiased estimate of
prediction ability in practice.

Key validation statistics:

| Statistic | Formula | Interpretation |
|-----------|---------|---------------|
| RMSEP | $\sqrt{\frac{1}{n}\sum(\hat{y}_i - y_i)^2}$ | Prediction error; same units as **y** |
| Bias | $\frac{1}{n}\sum(\hat{y}_i - y_i)$ | Systematic over/under-prediction |
| RPD | $\text{SD}(y_{\text{test}}) / \text{RMSEP}$ | Performance relative to population spread |
| RPIQ | $\text{IQR}(y_{\text{test}}) / \text{RMSEP}$ | As RPD but robust to outliers |
| R² | $1 - \frac{\sum(y_i - \hat{y}_i)^2}{\sum(y_i - \bar{y})^2}$ | Coefficient of determination |

---

## 6.2 Rebuild the calibration model

We reproduce the full workflow from Notebook 05 to have all objects available.

```{r rebuild-model}
data(NIRsoil)

spectra     <- NIRsoil$spc
wavelengths <- as.numeric(colnames(spectra))
spectra_snv <- standardNormalVariate(spectra)

keep <- !is.na(NIRsoil$Nt)
X    <- spectra_snv[keep, ]
y    <- NIRsoil$Nt[keep]

set.seed(42)
ks <- kenStone(X, k = round(nrow(X) * 0.7), metric = "mahal", pc = 10)

X_cal  <- X[ks$model, ]; y_cal  <- y[ks$model]
X_test <- X[ks$test,  ]; y_test <- y[ks$test]

pls_model <- plsr(
  y_cal ~ X_cal,
  ncomp      = 20,
  validation = "CV",
  segments   = 10,
  method     = "oscorespls"
)

# Select optimal ncomp
rmsecv    <- RMSEP(pls_model, estimate = "CV")$val[1, 1, ]
opt_ncomp <- (0:(length(rmsecv) - 1))[which.min(rmsecv)]
cat("Optimal LVs:", opt_ncomp, "\n")
```

---

## 6.3 Predictions on the test set

```{r test-predictions}
y_hat_test <- predict(pls_model, newdata = X_test, ncomp = opt_ncomp)[, , 1]
```

---

## 6.4 Validation statistics

```{r validation-stats}
val_stats <- function(y_obs, y_pred) {
  n    <- length(y_obs)
  res  <- y_pred - y_obs
  rmse <- sqrt(mean(res^2))
  bias <- mean(res)
  r2   <- 1 - sum(res^2) / sum((y_obs - mean(y_obs))^2)
  rpd  <- sd(y_obs) / rmse
  rpiq <- IQR(y_obs) / rmse
  data.frame(n = n, RMSE = rmse, Bias = bias, R2 = r2, RPD = rpd, RPIQ = rpiq)
}

# Calibration
y_hat_cal <- predict(pls_model, ncomp = opt_ncomp)[, , 1]
stats_cal  <- val_stats(y_cal, y_hat_cal)

# Cross-validation
y_hat_cv <- pls_model$validation$pred[, 1, opt_ncomp]
stats_cv  <- val_stats(y_cal, y_hat_cv)

# External test
stats_test <- val_stats(y_test, y_hat_test)

# Summary table
stats_all <- rbind(
  data.frame(Set = "Calibration",     stats_cal),
  data.frame(Set = "Cross-validation", stats_cv),
  data.frame(Set = "Test",            stats_test)
)

knitr::kable(stats_all, digits = 4,
             caption = paste0("Validation statistics — PLS (", opt_ncomp, " LVs), Nt (%)"))
```

### Interpreting RPD

| RPD range | Interpretation |
|-----------|---------------|
| < 1.5 | Model not useful for prediction |
| 1.5 – 2.0 | Rough screening only |
| 2.0 – 2.5 | Approximate quantitative prediction |
| 2.5 – 3.0 | Good quantitative prediction |
| > 3.0 | Excellent quantitative prediction |

*(Dunn et al., 2002; Williams & Norris, 2001)*

---

## 6.5 Predicted vs measured plots

```{r pred-vs-meas}
# Combine calibration and test data for a joint plot
plot_df <- bind_rows(
  data.frame(measured = y_cal,  predicted = y_hat_cal,
             set = "Calibration"),
  data.frame(measured = y_test, predicted = y_hat_test,
             set = "Test")
)

ggplot(plot_df, aes(x = measured, y = predicted, colour = set)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_colour_manual(values = c("Calibration" = "#2166AC",
                                  "Test"        = "tomato")) +
  labs(
    title  = paste0("Predicted vs measured Nt — PLS (", opt_ncomp, " LVs)"),
    x      = "Measured Nt (%)",
    y      = "Predicted Nt (%)",
    colour = NULL
  ) +
  coord_equal() +
  theme_bw() +
  theme(legend.position = "bottom")
```

---

## 6.6 Residual diagnostics

```{r residual-plot}
resid_df <- data.frame(
  measured   = y_test,
  predicted  = y_hat_test,
  residual   = y_hat_test - y_test
)

p_hist <- ggplot(resid_df, aes(x = residual)) +
  geom_histogram(bins = 20, fill = "#2166AC", colour = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "tomato") +
  labs(title = "Residual distribution",
       x = "Residual (predicted − measured)", y = "Count") +
  theme_bw()

p_vs <- ggplot(resid_df, aes(x = predicted, y = residual)) +
  geom_point(alpha = 0.6, colour = "#2166AC") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "tomato") +
  labs(title = "Residuals vs fitted",
       x = "Predicted Nt (%)", y = "Residual") +
  theme_bw()

p_hist + p_vs
```

A well-behaved model should show:
* Residuals roughly symmetric around zero (no systematic bias).
* No funnel pattern (homoscedastic errors).
* No trend in residuals vs fitted values.

---

## 6.7 Leverage and influence

Samples with high **leverage** (far from the centroid in X-space) may
disproportionately influence the model.  `pls` provides the hat-values
(diagonal of the hat matrix) to assess leverage.

```{r leverage}
hat_cal  <- hatvalues(pls_model, ncomp = opt_ncomp)
infl_df  <- data.frame(
  sample    = seq_along(hat_cal),
  leverage  = hat_cal,
  residual  = y_hat_cal - y_cal
)

ggplot(infl_df, aes(x = leverage, y = residual)) +
  geom_point(alpha = 0.6, colour = "#2166AC") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 2 * (opt_ncomp + 1) / nrow(X_cal),
             linetype = "dashed", colour = "tomato") +
  labs(
    title    = "Leverage vs residuals (calibration)",
    subtitle = "Dashed vertical line: leverage threshold = 2(k+1)/n",
    x = "Leverage (hat value)",
    y = "Calibration residual"
  ) +
  theme_bw()
```

---

## 6.8 Reporting model performance

When reporting a NIR calibration model, always include:

1. **Pre-processing** applied and parameters used.
2. **Number of latent variables** and selection criterion.
3. **Calibration statistics**: n, RMSEC, R²_c.
4. **Cross-validation statistics**: RMSECV, R²_cv.
5. **External validation statistics**: RMSEP, Bias, R²_p, RPD, RPIQ, n_test.
6. **Scatter plot** of predicted vs measured for both sets.
7. **Population description**: mean, SD, and range of the reference property.

---

## 6.9 Summary

In this final notebook we:

* Evaluated the PLS model on an independent test set.
* Computed RMSEP, Bias, R², RPD and RPIQ.
* Examined residuals for systematic errors.
* Assessed sample leverage and influence.

### Next steps

* Explore **local / memory-based** modelling with `resemble::mbl()` for
  heterogeneous datasets.
* Investigate **variable selection** methods (iPLS, VIP, competitive adaptive
  reweighted sampling).
* Apply the calibration model to **new handheld measurements** collected in
  the field.

---

## Session information

```{r session-info}
sessionInfo()
```
